---
title: "Assignment 3. part a. Arisov DATA 607"
author: "Maxim Arisov"
format: html
date: 09/20/26
code-fold: true
toc: true
editor: visual
---

## **Global Baseline Estimates**

## **Approach**

For this assignment, I will use the movie ratings data XLSX collected from the class survey to create a Global Baseline Estimate recommendation system in R studio.

First, I will organize the movie ratings and check for missing ratings.

Second, I will calculate the average rating for each user and compare it with the overall average rating.

Third, I will calculate the average rating for each movie and compare it with the overall average rating.

After above steps, I will use the Global Baseline Estimate formula to predict a missing movie rating. The prediction will use the overall average, the user's rating difference from the average, and the movie's rating difference from the average.

Lastly, I will use the predicted rating to create a movie recommendation and explain the results.

GitHub link:

<https://github.com/Maximus2485/Assignment-3.-part-a.-Global-Baseline-Estimates.-/blob/main/MovieRatings.xlsx>

## Data Source

The 3 datasets were provided by professor for this assignment. I uploaded the dataset to my GitHub repository so I could access and use the data in RStudio.

global baseline

<https://raw.githubusercontent.com/Maximus2485/Assignment-3.-part-a.-Global-Baseline-Estimates.-/refs/heads/main/MovieRatings.global%20baseline.csv>

mean centered

<https://raw.githubusercontent.com/Maximus2485/Assignment-3.-part-a.-Global-Baseline-Estimates.-/refs/heads/main/MovieRatings.mean%20centered.csv>

problem statement

<https://raw.githubusercontent.com/Maximus2485/Assignment-3.-part-a.-Global-Baseline-Estimates.-/refs/heads/main/MovieRatings.problem%20statement.csv>

## Code Base

**Lets load the three datasets**

```{r}
# Problem statement data
problem <- read.csv(
  "https://raw.githubusercontent.com/Maximus2485/Assignment-3.-part-a.-Global-Baseline-Estimates.-/refs/heads/main/MovieRatings.problem%20statement.csv",
  header = FALSE
)

# Mean-centered data
mean_centered <- read.csv(
  "https://raw.githubusercontent.com/Maximus2485/Assignment-3.-part-a.-Global-Baseline-Estimates.-/refs/heads/main/MovieRatings.mean%20centered.csv"
)

# Global baseline data
baseline <- read.csv(
  "https://raw.githubusercontent.com/Maximus2485/Assignment-3.-part-a.-Global-Baseline-Estimates.-/refs/heads/main/MovieRatings.global%20baseline.csv"
)

head(problem)
head(mean_centered)
head(baseline)
```

**Processing of data before analysis**

```{r}
#Because the problem file has no column names, lets give them names.
names(problem) <- c(
  "Critic",
  "CaptainAmerica",
  "Deadpool",
  "Frozen",
  "JungleBook",
  "PitchPerfect2",
  "StarWarsForce"
)

problem

#Param's PitchPerfect2 value is "?" in the professor's data. Lets change that to a missing value.
problem[problem == "?"] <- NA

#Lets convert the movie columns to numbers
problem[2:7] <- lapply(problem[2:7], as.numeric)
```

**Lets calculate the overall mean**

```{r}
overall_mean <- mean(
  as.matrix(problem[2:7]),
  na.rm = TRUE
)

overall_mean
```

\# I got overall_mean 3.93, which matches professor's Global Baseline Estimates.

**Calculate each movie average**

```{r}
movie_avg <- colMeans(
  problem[2:7],
  na.rm = TRUE
)

movie_avg
```

I got move_ave , which match professor's value in Global Baseline Estimates.

**Calculate how each movie is from the overall mean:**

```{r}
movie_difference <- movie_avg - overall_mean

movie_difference
```

I got movie_difference , which match professor's value movie avg - mean movie in Global Baseline Estimates.

**Let calculate each user's average**

```{r}
user_avg <- rowMeans(
  problem[2:7],
  na.rm = TRUE
)

user_avg
```

I got user_avg, which match professor's value user avg in Global Baseline Estimates.

**Calculate each user's difference form the overall mean**

```{r}
user_difference <- user_avg - overall_mean

#Add these values to the dataset
problem$user_avg <- user_avg
problem$user_difference <- user_difference

problem
```

**Lets predict Param's rating**

```{r}
param_difference <- problem$user_difference[
  problem$Critic == "Param"
]

pitch_difference <- movie_difference["PitchPerfect2"]

predicted_rating <- overall_mean +
                    param_difference +
                    pitch_difference

predicted_rating
```

I got predicted_rating approximately 2.28, which reproduces the professor's calculation.

The overall average movie rating is about 3.93. Param's ratings are about 0.43 below the overall average, and Pitch Perfect 2 is about 1.22 below the overall average. Using the Global Baseline Estimate, Param's predicted rating for Pitch Perfect 2 is about 2.28. This means the model estimates that Param would give Pitch Perfect 2 a rating of about 2.3.

**Lets make a movie recommendation**

```{r}
#I calculated predicted ratings for the movies that each user did not rate. For each user, I selected the movie with the highest predicted rating as the recommendation. A higher predicted rating means the Global Baseline Estimate expects the user to give that movie a higher rating.

library(dplyr)
library(tidyr)

recommendations <- problem %>%
  pivot_longer(
    CaptainAmerica:StarWarsForce,
    names_to = "Movie",
    values_to = "Rating"
  ) %>%
  filter(is.na(Rating)) %>%
  mutate(
    Predicted_Rating =
      overall_mean +
      user_difference +
      movie_difference[Movie]
  ) %>%
  group_by(Critic) %>%
  slice_max(Predicted_Rating, n = 1) %>%
  select(Critic, Movie, Predicted_Rating)

recommendations
```

## Conclusion

In this assignment, I used the Global Baseline Estimate to predict missing movie ratings and make movie recommendations. I first calculated the overall average rating, each user's difference from the overall average, and each movie's difference from the overall average. I then used these values to calculate predicted ratings for movies that each user had not rated.

For each user, I selected the movie with the highest predicted rating as the recommendation. For example, the model recommended CaptainAmerica to Dan with a predicted rating of about 5.34.

Overall, the Global Baseline Estimate provides a simple way to predict missing ratings and recommend movies based on both the user's rating behavior and the general ratings of each movie.

## References

Google Deep Mind. (2026). Gemini 3 Flash \[Large language model\]. \[<https://gemini.google.com.%5D(https://gemini.google.com/)> Accessed September 20th, 2026.

Wickham, H., Çetinkaya-Rundel, M., & Grolemund, G. (2023). ***R for data science*** (2nd ed.). O'Reilly Media. <https://r4ds.hadley.nz/>
